---
title: "Asthma among public housing residents in King County"
subtitle: "Data slides"
header-includes:
  - \usepackage{booktabs}
  - \usepackage{makecell}
  - \usepackage{multirow}
output: 
  powerpoint_presentation:
    reference_doc: "C:/Users/mathesal/King County/Laurent, Amy - DASHH-External/DASHH2.0/Convening/DASHH convening 2019-02 - data slides template.pptx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(knitr.kable.NA = '')

# Function to format cell
bold_cell <- function(x) {
  cell_spec(x, "latex", bold = ifelse(as.numeric(str_replace(x, "<", "")) < 0.05, T, F))
}

```

## How many people have asthma and who are they?

### CCW asthma increases with age and is highest among HCV recipients
```{r ccw_age_sub, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
ccw_asthma_age_sub_g + facet_wrap( ~ year_text, ncol = 1)
```

### HEDIS asthma also increases with age (2018, non-dual Medicaid members only)
```{r hedis_age_sub, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
hedis_asthma_all_age_sub_g + facet_wrap( ~ amr_type, ncol = 1)
```

### Gap between non-PHA and HCV/PH highest for Latinos and whites (CCW asthma) 
```{r ccw_race_sub, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
ccw_asthma_race_sub_18 %>%
  ggplot(aes(fill = pha_subsidy, y = prop, x = race_eth_me)) + 
  geom_bar(position = "dodge", stat = "identity", color = "#333333") +
  scale_fill_manual(values = c("#79706e", "#2c7bb6", "#87d180"), name = "PHA subsidy type") + 
  xlab("Race/ethnicity") + ylab("Number per 1,000") + 
  ggtitle("2018, non-dual Medicaid members") +
  theme_ipsum_ps() +
  theme(panel.grid.major.y = element_line(color = "grey90"),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor = element_blank())
```

### People receiving HCVs are most likely to have HEDIS-defined asthma
Adjusted for age, gender, and race/ethnicity:
* People receiving HCVs were `r abs(round((asthma_m_1yr_sub_coef_hcv$estimate-1)*100, 1))`% 
`r ifelse(asthma_m_1yr_sub_coef_hcv$estimate > 1, "more", "less")` likely to meet the one-year HEDIS definition for asthma 
than non-PHA Medicaid recipients 
**(`r ifelse(asthma_m_1yr_sub_coef_hcv$p.value < 0.05, "significantly different", "not significant")`)**
* People living in public housing were `r abs(round((asthma_m_1yr_sub_coef_ph$estimate-1)*100, 1))`% 
`r ifelse(asthma_m_1yr_sub_coef_ph$estimate > 1, "more", "less")` likely to meet the one-year HEDIS definition for asthma 
than non-PHA Medicaid recipients 
**(`r ifelse(asthma_m_1yr_sub_coef_ph$p.value < 0.05, "significantly different", "not significant")`)**


### If the level of asthma among PHA recipients looked like non-PHA Medicaid members...
* **`r ccw_asthma_all_demog_18_avoidable$difference[ccw_asthma_all_demog_18_avoidable$pha_subsidy == "HCV"]`** 
fewer people with asthma in the HCV group
* **`r ccw_asthma_all_demog_18_avoidable$difference[ccw_asthma_all_demog_18_avoidable$pha_subsidy == "PH"]`** 
fewer people with asthma in the PH group


### Strong north-south gradient in percent with CCW asthma (King County Medicaid and Medicare members, by ZIP, 2016)
```{r map_16_prop, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
geo_zip_map_16_prop
```

### Additional hot spots of CCW asthma appear when looking at King County non-dual Medicaid members (by ZIP, 2018)
```{r map_18_prop, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
geo_zip_map_18_prop
```

### Adjusting for demographics shows a similar pattern (risk of CCW asthma among Medicaid and Medicare members, by ZIP, 2016)
```{r map_16_inla, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
geo_zip_map_16_inla
```


## How well do people manage their asthma?
### Younger children are more likely to experience ED visits primarily due to asthma
```{r acute_18_sub, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
acute_18_age_sub_g
```

### Adults are more likely to experience ED visits and hospitalizations where asthma is a contributor
```{r acute_any_18_sub, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
acute_18_all_dx_age_sub_g + 
  facet_wrap( ~ type, ncol = 1, scales = "free_y")
```

### Little difference in managing asthma across ages and PHA subsidy type
```{r hedis_amr_age_sub, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
hedis_amr_age_1yr_sub %>%
  ggplot(aes(fill = pha_subsidy, y = prop, x = age_yr_asthma)) + 
  geom_bar(position = "dodge", stat = "identity", color = "#333333") +
  scale_fill_manual(values = c("#79706e", "#2c7bb6", "#87d180"), name = "Agency") + 
  xlab("Age (years)") + ylab("Percent") + 
  ggtitle("Percent who meet the AMR (one-year HEDIS asthma, 2018, non-dual Medicaid members)") +
  theme_ipsum_ps() +
  theme(panel.grid.major.y = element_line(color = "grey90"),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor = element_blank())
```


### HCV recipients and PH residents manage asthma at the same level as non-PHA Medicaid members
Adjusted for age, gender, and race/ethnicity:
* People receiving HCVs were `r abs(round((amr_m_1yr_sub_coef_hcv$estimate-1)*100, 1))`% 
`r ifelse(amr_m_1yr_sub_coef_hcv$estimate > 1, "more", "less")` likely to meet the one-year HEDIS definition for asthma 
than non-PHA Medicaid recipients 
**(`r ifelse(amr_m_1yr_sub_coef_hcv$p.value < 0.05, "significantly different", "not significant")`)**
* People living in public housing were `r abs(round((amr_m_1yr_sub_coef_ph$estimate-1)*100, 1))`% 
`r ifelse(amr_m_1yr_sub_coef_ph$estimate > 1, "more", "less")` likely to meet the one-year HEDIS definition for asthma 
than non-PHA Medicaid recipients 
**(`r ifelse(amr_m_1yr_sub_coef_ph$p.value < 0.05, "significantly different", "not significant")`)**

::: The same holds true after adjusting for age, gender, and race/ethnicity :::



## Extra slides
### CCW asthma increases with age and is highest among HCV recipients
```{r ccw_age_sub_ci, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
ccw_asthma_age_sub_g + 
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.2,
                position = position_dodge(width = 0.9)) +
  facet_wrap( ~ year_text, ncol = 1)
```


### CCW asthma increases with age and is higher among PHA recipients
```{r ccw_age_pha_ci, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
ccw_asthma_age_pha_g + 
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.2,
                position = position_dodge(width = 0.9)) +
  facet_wrap( ~ year_text, ncol = 1)
```


### HEDIS asthma also increases with age group (2018, non-dual Medicaid members only)
```{r hedis_age_pha, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
hedis_asthma_all_age_sub_g + 
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.2,
                position = position_dodge(width = 0.9)) +
  facet_wrap( ~ amr_type, ncol = 1)
```


### Gap between non-PHA and HCV/PH highest for Latinos and whites (CCW asthma) 
```{r ccw_race_sub_ci, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
ccw_asthma_race_sub_g + 
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.2,
                position = position_dodge(width = 0.9)) +
  facet_wrap( ~ year_text, ncol = 1)
```


## Risk of having asthma after adjusting for age, gender, and race/ethnicity
### Risk of having CCW asthma among non-dual Medicaid members in King County, by ZIP (2018)
```{r map_18_inla, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
geo_zip_map_18_inla
```

### Risk of having CCW asthma among Medicaid and Medicare members in King County, by ZIP (2016)
```{r map_16_inla_pha, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
geo_zip_map_16_inla + geom_sf(data = portfolios) + 
  geom_label_repel(data = portfolios, aes(x = Lon, y = Lat, label = Portfolio),
                   nudge_x = c(-0.2, 0.2, -0.5, 0.3, -0.5, -0.3, -0.3, -0.5, 0.1),
                   nudge_y = c(0, 0, 0, 0.05, 0, 0, 0.01, 0.03, -0.03))
```

### Risk of having CCW asthma among non-dual Medicaid members in King County, by ZIP (2018)
```{r map_18_inla_pha, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
geo_zip_map_18_inla + geom_sf(data = portfolios) + 
  geom_label_repel(data = portfolios, aes(x = Lon, y = Lat, label = Portfolio),
                   nudge_x = c(-0.2, 0.2, -0.5, 0.3, -0.5, -0.3, -0.3, -0.5, 0.1),
                   nudge_y = c(0, 0, 0, 0.05, 0, 0, 0.01, 0.03, -0.03))
```

### Adjusted HEDIS-defined asthma results (PHA vs. non-PHA)
After adjusting for age, gender, and race/ethnicity, people receiving PHA support were:
- `r abs(round((asthma_m_pha_coef_pha$estimate-1)*100, 1))`% 
`r ifelse(asthma_m_pha_coef_pha$estimate > 1, "more", "less")` likely to meet the HEDIS definition for persistent asthma 
(`r ifelse(asthma_m_pha_coef_pha$p.value < 0.05, "significantly different", "not significant")`)
- `r abs(round((asthma_m_1yr_pha_coef_pha$estimate-1)*100, 1))`% 
`r ifelse(asthma_m_1yr_pha_coef_pha$estimate > 1, "more", "less")` likely to meet the one-year HEDIS definition for asthma 
(`r ifelse(asthma_m_1yr_pha_coef_pha$p.value < 0.05, "significantly different", "not significant")`)
- `r abs(round((amr_m_pha_coef_pha$estimate-1)*100, 1))`% 
`r ifelse(amr_m_pha_coef_pha$estimate > 1, "more", "less")` likely to meet the desired AMR (persistent asthma) 
(`r ifelse(amr_m_pha_coef_pha$p.value < 0.05, "significantly different", "not significant")`)
- `r abs(round((amr_m_1yr_pha_coef_pha$estimate-1)*100, 1))`% 
`r ifelse(amr_m_1yr_pha_coef_pha$estimate > 1, "more", "less")` likely to meet the desired AMR (one-year asthma) 
(`r ifelse(amr_m_1yr_pha_coef_pha$p.value < 0.05, "significantly different", "not significant")`)

## ED visits/hospitalizations comparing PHA subsidy types (with CIs)
###
```{r acute_18_sub_ci, echo=FALSE, warning=FALSE, fig.width=10, fig.height=7}
acute_18_age_sub_g +
    geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.2,
                position = position_dodge(width = 0.9))
```

###
```{r acute_any_18_sub_ci, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
acute_18_all_dx_age_sub_g + 
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.2,
                position = position_dodge(width = 0.9)) + 
  facet_wrap( ~ type, ncol = 1, scales = "free_y")
```

## ED visits/hospitalizations comparing KCHA and SHA (with CIs)
###
```{r acute_18_pha_ci, echo=FALSE, warning=FALSE, fig.width=10, fig.height=7}
acute_18_age_pha_g + 
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.2,
                position = position_dodge(width = 0.9))
```

###
```{r acute_any_18_pha_ci, echo=FALSE, warning=FALSE, fig.width=11, fig.height=7}
acute_18_all_dx_age_pha_g + 
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.2,
                position = position_dodge(width = 0.9)) + 
  facet_wrap( ~ type, ncol = 1, scales = "free_y")
```
