---
title: "Asthma among public housing residents in King County"
header-includes:
  - \usepackage{booktabs}
  - \usepackage{makecell}
  - \usepackage{multirow}
output: 
  powerpoint_presentation:
    reference_doc: "C:/Users/mathesal/King County/Laurent, Amy - DASHH-External/DASHH2.0/Convening/DASHH convening 2019-02 - overall slides.pptx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(knitr.kable.NA = '')

# Function to format cell
bold_cell <- function(x) {
  cell_spec(x, "latex", bold = ifelse(as.numeric(str_replace(x, "<", "")) < 0.05, T, F))
}

```



# CCW asthma increases with age and is higher among PHA recipients
```{r ccw_age, echo=FALSE, warning=FALSE, fig.width=9, fig.height=6}
ccw_asthma_age_g + facet_wrap( ~ year_text, ncol = 1)
```

# HEDIS asthma by age group (2018, non-dual Medicaid members only)
```{r hedis_age, echo=FALSE, warning=FALSE, fig.width=9, fig.height=6}
hedis_asthma_all_age_g + facet_wrap( ~ amr_type, ncol = 1)
```

# Adjusted HEDIS-defined asthma results
After adjusting for age, gender, and race/ethnicity, people receiving PHA support were:

- `r abs(round((asthma_m_coef_pha$estimate-1)*100, 1))`% 
`r ifelse(asthma_m_coef_pha$estimate > 1, "more", "less")` likely to meet the HEDIS definition for persistent asthma 
(`r ifelse(asthma_m_coef_pha$p.value < 0.05, "significantly different", "not significant")`)
- `r abs(round((asthma_m_1yr_coef_pha$estimate-1)*100, 1))`% 
`r ifelse(asthma_m_1yr_coef_pha$estimate > 1, "more", "less")` likely to meet the one-year HEDIS definition for asthma 
(`r ifelse(asthma_m_1yr_coef_pha$p.value < 0.05, "significantly different", "not significant")`)
- `r abs(round((amr_m_coef_pha$estimate-1)*100, 1))`% 
`r ifelse(amr_m_coef_pha$estimate > 1, "more", "less")` likely to meet the desired AMR (persistent asthma) 
(`r ifelse(amr_m_coef_pha$p.value < 0.05, "significantly different", "not significant")`)
- `r abs(round((amr_m_1yr_coef_pha$estimate-1)*100, 1))`% 
`r ifelse(amr_m_1yr_coef_pha$estimate > 1, "more", "less")` likely to meet the desired AMR (one-year asthma) 
(`r ifelse(amr_m_1yr_coef_pha$p.value < 0.05, "significantly different", "not significant")`)


# Risk of having CCW asthma among Medicaid and Medicare members in King County, by ZIP (2016)
```{r map_16, echo=FALSE, warning=FALSE, fig.width=9, fig.height=6}
geo_zip_map(geo_zip_16, title = F)
```

# Risk of having CCW asthma among Medicaid (non-dual) members in King County, by ZIP (2018)
```{r map_18, echo=FALSE, warning=FALSE, fig.width=9, fig.height=6}
geo_zip_map(geo_zip_18, title = F)
```



