---
title: "Asthma among public housing residents"
author: "Alastair Matheson"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
  - \usepackage{booktabs}
  - \usepackage{makecell}
  - \usepackage{multirow}
output: 
  powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(knitr.kable.NA = '')

# Function to format cell
bold_cell <- function(x) {
  cell_spec(x, "latex", bold = ifelse(as.numeric(str_replace(x, "<", "")) < 0.05, T, F))
}

```



## CCW asthma by age group
```{r ccw_age}
ccw_asthma_age_g + facet_wrap( ~ year_text, ncol = 1)

```



## Asthma and AMR regression

The tables below shows output from some basic binomial regressions looking at 1-year asthma and asthma medication ratio (AMR).

For both asthma and AMR, three regressions were run:
1) Comparing PHA vs. non-PHA (adjusting for age, race/ethnicity, and gender)
2) Comparing KCHA vs. SHA among hard unit residents (adjusting for age, race/ethnicity, gender, length of time in housing, and operator type)
3) Comparing KCHA vs. SHA among soft unit residents (adjusting for age, race/ethnicity, gender, length of time in housing, and voucher type)

Note that the term 'have asthma' is used below. This should be understood to mean met the HEDIS definition for asthma, which means that 
the person must have sought care for this condition and received an asthma diagnosis.

When interpreting results, differential access to healthcare as a driver of differences should be kept in mind (even though everyone included 
was enrolled in Medicaid for at least 11 months in 2017).



## Asthma comparing PHA/non-PHA
After adjusting for age, gender, and race/ethnicity, non-dual PHA residents on Medicaid were 48% (95% CI: 39-58%) more likely 
to meet the definition of asthma than non-PHA Medicaid recipients.

For Medicaid recipients overall:
- Age, gender, and race/ethnicity were all significant overall.
- The odds of having asthma increased with age (those aged 51-64 were 48% (95% CI: 36-60%) more likely to have asthma than those aged 5-11).
- Males were 18% less likely to have an asthma diagnosis than females (95% CI: 14-22%)

```{r table.asthma.pha.non, echo = F}
out1 <- model_output_overall %>% 
  filter(model == "Asthma comparing PHA/non-PHA") %>%
  select(-model) %>%
  mutate(p = case_when(!is.na(p) ~ bold_cell(p)),
         chi2 = case_when(!is.na(chi2) ~ bold_cell(chi2))) %>%
  kable(format = "latex", booktabs = T, escape = F,
        caption = paste0("Asthma comparing PHA and non-PHA (n = ", nobs(asthma_overall), ")")) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  footnote(general = "chi2 column shows the output of the Wald test for the entire group") %>%
  kable_styling(latex_options = "hold_position")

out1 <- gsub("(_)", "\\\\\\1", out1)
out1

```

\newpage

## AMR comparing PHA/non-PHA
After adjusting for age, gender, and race/ethnicity, PHA residents with asthma were as likely to manage their asthma as non-PHA Medicaid
recipients.

For Medicaid recipients overall:
- Those who identified as American Indian/Alaskan Native were least likely to manage their asthma.
- Asthma was most likely to be managed among younger children (5-11 years), though this was only significant when compared with youth
aged 12-18 and adults aged 19-30.
Males were slightly less likely to manage their astham (~10%, 95% CI: 0-18%, p = 0.05)


```{r table.amr.pha.non, echo = F}
out2 <- model_output_overall %>% 
  filter(model == "AMR comparing PHA/non-PHA") %>%
  select(-model) %>%
  mutate(p = case_when(!is.na(p) ~ bold_cell(p)),
         chi2 = case_when(!is.na(chi2) ~ bold_cell(chi2))) %>%
  kable(format = "latex", booktabs = T,  escape = F,
        caption = paste0("AMR comparing PHA and non-PHA (n = ", nobs(amr_overall), ")")) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  footnote(general = "chi2 column shows the output of the Wald test for the entire group") %>%
  kable_styling(latex_options = "hold_position")

out2 <- gsub("(_)", "\\\\\\1", out2)
out2

```

\newpage

## Asthma in PHA hard units
After adjusting for age, gender, race/ethnicity, length of time in housing, and operator type, there was no significant difference 
in the odds of having asthma between SHA and KCHA residents in hard units.


```{r table.asthma.hard, echo = F}
out3 <- model_output_overall %>% 
  filter(model == "Asthma in PHA hard units") %>%
  select(-model) %>%
  mutate(p = case_when(!is.na(p) ~ bold_cell(p)),
         chi2 = case_when(!is.na(chi2) ~ bold_cell(chi2))) %>%
  kable(format = "latex", booktabs = T,  escape = F,
        caption = paste0("Asthma in PHA hard units (n = ", nobs(asthma_pha_hard), ")")) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  footnote(general = "chi2 column shows the output of the Wald test for the entire group") %>%
  kable_styling(latex_options = "hold_position")

out3 <- gsub("(_)", "\\\\\\1", out3)
out3

```

\newpage

## AMR in PHA hard units
After adjusting for age, gender, race/ethnicity, length of time in housing, and operator type, there was no significant difference 
in asthma management between SHA and KCHA residents in hard units.

```{r table.amr.hard, echo = F}
out4 <- model_output_overall %>% 
  filter(model == "AMR in PHA hard units") %>%
  select(-model) %>%
  mutate(p = case_when(!is.na(p) ~ bold_cell(p)),
         chi2 = case_when(!is.na(chi2) ~ bold_cell(chi2))) %>%
  kable(format = "latex", booktabs = T, escape = F,
        caption = paste0("AMR in PHA hard units (n = ", nobs(amr_pha_hard), ")")) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  footnote(general = "chi2 column shows the output of the Wald test for the entire group") %>%
  kable_styling(latex_options = "hold_position")

out4 <- gsub("(_)", "\\\\\\1", out4)
out4
```


\newpage

## Asthma in PHA soft units
After adjusting for age, gender, race/ethnicity, length of time in housing, and voucher type, there was no significant difference 
in the odds of having asthma between SHA and KCHA residents in soft units.

```{r table.asthma.soft, echo = F}
out5 <- model_output_overall %>% 
  filter(model == "Asthma in PHA soft units") %>%
  select(-model) %>%
  mutate(p = case_when(!is.na(p) ~ bold_cell(p)),
         chi2 = case_when(!is.na(chi2) ~ bold_cell(chi2))) %>%
  kable(format = "latex", booktabs = T, escape = F,
        caption = paste0("Asthma in PHA soft units (n = ", nobs(asthma_pha_soft), ")")) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  footnote(general = "chi2 column shows the output of the Wald test for the entire group") %>%
  kable_styling(latex_options = "hold_position")

out5 <- gsub("(_)", "\\\\\\1", out5)
out5
```

\newpage

## AMR in PHA soft units
After adjusting for age, gender, race/ethnicity, length of time in housing, and voucher type, there was no significant difference 
in asthma management between SHA and KCHA residents in soft units.


```{r table.amr.soft, echo = F}
out6 <- model_output_overall %>% 
  filter(model == "AMR in PHA soft units") %>%
  select(-model) %>%
  mutate(p = case_when(!is.na(p) ~ bold_cell(p)),
         chi2 = case_when(!is.na(chi2) ~ bold_cell(chi2))) %>%
  kable(format = "latex", booktabs = T, escape = F,
        caption = paste0("AMR in PHA soft units (n = ", nobs(amr_pha_soft), ")")) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  footnote(general = "chi2 column shows the output of the Wald test for the entire group") %>%
  kable_styling(latex_options = "hold_position")

out6 <- gsub("(_)", "\\\\\\1", out6)
out6
```
